The need for accuracy and smoothness in numerical simulations

### 2024-08-30 slides/ostrava/cckm-ppam-2024.tex
Added more pictures to the slides for PPAM 2024

\graphicspath{{figures/}{data/}}

@@

  The system of differential algebraic equations
  \begin{align}
  \bm{q}'(t) &= \bm{v}(t), \\
  \bm{M}\bm{v}'(t) &= \bm{f}(\bm{q}(t)) - \bm{G}(\bm{q}(t))^T\bm{\lambda}(t), \\
  \bm{g}(\bm{q}(t)) &= \bm{0}. 
  \end{align}
  is solved using the SHAKE algorithm
  \begin{align}
  \bm{v}_{n+1/2} &= \bm{v}_{n-1/2} + \bm{h} \bm{M}^{-1} \left( \bm{f}(\bm{q}_n) - \bm{G}(\bm{q}_n)^T \bm{\lambda}_n \right), \\
  \bm{q}_{n+1} &= \bm{q}_n + h \bm{v}_{n + 1/2}, \\
  \bm{g}(\bm{q}_{n+1}) &= \bm{0}. \label{equ:constraint}
\end{align}
}

@@

  It might be possible to assert experimentally if
  \vspace{0.5cm}
  \begin{enumerate} \setlength\itemsep{1em}
  \item your functions are sufficiently smooth
  \item your equations are solved with sufficient accuracy
  \end{enumerate}
}

@@

Define 
\begin{equation}
  R_h := \frac{A_h - A_{2h}}{2^p - 1}, \quad F_h:= \frac{A_{2h} - A_{4h}}{A_h - A_{2h}} 
\end{equation}
If 
\begin{equation}
  E_h = T - A_h = \alpha h^p + \beta h^q + O(h^r), \quad h \rightarrow 0_+
\end{equation}
then
\begin{equation}
  \frac{E_h - R_h}{h^q} \rightarrow \text{constant}
\end{equation}
and
\begin{equation}
  \frac{2^p - F_h}{h^{q-p}} \rightarrow \text{constant}
\end{equation}
Why is this significant when the computer returns $\hat{A}_h$ and $$\hat{A}_h \not = A_h$$
}

@@

\frame{\frametitle{What can kill an AEX?}

>>>>

\graphicspath{{../../report/fig/}}

@@

  The system of differential algebraic equations
  \begin{align}
  \bm{q}'(t) &= \bm{v}(t), \\
  \bm{M}\bm{v}'(t) &= \bm{f}(\bm{q}(t)) - \bm{G}(\bm{q}(t))^T\bm{\lambda}(t), \\
  \bm{g}(\bm{q}(t)) &= \bm{0}. 
  \end{align}
  is solved using the SHAKE algorithm
  \begin{align}
  \bm{v}_{n+1/2} &= \bm{v}_{n-1/2} + \bm{h} \bm{M}^{-1} \left( \bm{f}(\bm{q}_n) - \bm{G}(\bm{q}_n)^T \bm{\lambda} \right), \\
  \bm{q}_{n+1} &= \bm{q}_n + h \bm{v}_{n + 1/2}, \\
  \bm{g}(\bm{q}_{n+1}) &= \bm{0}. \label{equ:constraint}
\end{align}
}

@@

  It might be possible to assert experimentally if
  \vspace{0.5cm}
  \begin{enumerate} \setlength\itemsep{1em}
  \item rounding errors are irrelevant
  \item your functions are sufficiently smooth
  \item your equations are solved with sufficient accuracy
  \end{enumerate}
}

@@

Define 
\begin{equation}
  R_h := \frac{A_h - A_{2h}}{2^p - 1}, \quad F_h:= \frac{A_{2h} - A_{4h}}{A_h - A_{2h}} 
\end{equation}
If 
\begin{equation}
  E_h = T - A_h = \alpha h^p + \beta h^q + O(h^r), \quad h \rightarrow 0_+
\end{equation}
then
\begin{equation}
  \frac{E_h - R_h}{h^q} \rightarrow \text{constant}
\end{equation}
and
\begin{equation}
  \frac{2^p - F_h}{h^{q-p}} \rightarrow \text{constant}
\end{equation}
Why is this significant? The computer returns $$\hat{A}_h \not = A_h$$
}

@@

\frame{\frametitle{When do we have an AEX?}

  Deriving an AEX is an exercise in Taylor expansions. If
  \begin{center}
    our functions are not many times differentiable
  \end{center}
  then the foundation crumbles.
  \vspace{1cm}
    \newline
  Every AEX refers to the exact value of the $A_h$. If
  $$\hat{A}_h \approx A_h$$
  is not a good approximation, then
  \begin{center}
    $\hat{A}_h$ will not behave nicely
  \end{center}
}

\frame{\frametitle{External ballistics: Accurate event location}

  \begin{figure}
    \includegraphics[width=10cm]{maxrange_rk1_tol53.pdf}
    \caption{RK1: The time to impact is computed with tolerance $\tau = 2^{-53}$. We can identify the asymptotic range and we trust the error estimate.}
  \end{figure}
}

\frame{\frametitle{External ballistics: Inaccurate event location}

  \begin{figure}
    \includegraphics[width=10cm]{maxrange_rk2_tol11.pdf}
    \caption{RK2: The time to impact is computed with tolerance $\tau = 2^{-11}$ and we cannot locate the asymptotic range nor trust the error estimate.}
  \end{figure}
}

\frame{\frametitle{External ballistics: Sufficiently event location}

  \begin{figure}
    \includegraphics[width=10cm]{maxrange_rk2_tol25.pdf}
    \caption{RK2: The time to impact is computed with tolerance $\tau = 2^{-25}$ and the asymptotic range begins to emerge}
  \end{figure}
}

\frame{\frametitle{Molecular dynamics: Truncation of force-fields}

  \begin{itemize}
  \item Simple simulation with 4 ions
  \item Each ion is attached to the origin with a spring
  \item The electrostatic forcefields are trunctated.
  \end{itemize}
}
